The brightness and spatial distributions of terrestrial radio sources

### 2013-04-05 articles/Article-RFI-distribution/article-rfi-distribution.tex
Todays work on the paper

Typical radio observations record a few hours of data, and the results are integrated. In these cases, excising only the interference that is apparent and thus above the noise often suffices, i.e., the observation can still reach the thermal noise limit of the instrument. A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, weak interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be persistently present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for the 21-cm Epoch of Reionisation (EoR) experiments, because they involve long integration times. Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}. Ground-based Cosmic Microwave Background (CMB) experiments are another class of experiments involving long integration times (e.g., \citealt{ska-cmb-subrahmanyan}). For these experiments, it is important to know the possible effect of low-level interference on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, van Haarlem et al. 2013, \textit{submitted}).
% TODO correct reference if van Haarlem et al is published

@@

\begin{figure}
\begin{center}\hspace{-5mm}\includegraphics[width=8cm]{img/plot-rayleigh-and-rfi-trimmed}
\caption{The Rayleigh and power-law distributions in a log-log plot. The power-law distribution (Eq.~\eqref{eq:two-dimensional-distribution}) has a constant slope of -2. Its integral is not equal to 1 in this example, because some form of flattening would be required to make this possible. The slope of the Rayleigh distribution in the limit of the origin is 1. Its maximum occurs where the amplitude value equals its mode $\sigma$, which is 1 in this example. For higher amplitudes, its slope decreases exponentially.}
\label{fig:rayleigh-and-rfi-distributions}
\end{center}
\end{figure}

@@

The stochastic nature of $I$, that is caused by the spread of transmitters with different intrinsic strengths, might also have an effect on the $\log N$ -- $\log S$ histograms. It is unlikely that $I$ follows a uniform or Gaussian distribution, because the distribution will contain few strong transmitters (such as radio stations) and many weak transmitters (such as remote controls). Therefore, variable $I$ might follow a power-law distribution by itself. It is likely that strong transmitters transmit more on average, and therefore contaminate more samples as well. High-power transmitters, such as radio stations, have a typical transmitting power in the order of 100 kW. Low-power transmitters, such as remote controls, transmit with an order of 100 mW or even less. Therefore, these devices have a spread of around 6 orders of magnitudes in power. As long as the exponent of the power law of $I$ is less steep (i.e., less negative) compared to the power law caused by the spatial distribution, the spatial distribution will dominate the histogram at high amplitudes. With a spatial $-1.5$ power law and the given transmitting powers, the low-power transmitters should contaminate a factor of $10^9$ more samples compared to the high-power transmitters to dominate the high-amplitude distribution, which is unlikely. Therefore, it is likely that the spatial distribution will dominate the power law in the histogram. Otherwise, a turn-over point should be visible in the histogram.

@@

\begin{table*}
\centering
\begin{minipage}{12cm}
\caption{Estimated distribution quantities per data set. }\label{table:dist-data-quantities}
\begin{tabular}{@{}llrr@{}}
\textbf{Symbol} & \textbf{Description} & \textbf{LBA set}& \textbf{HBA set} \\
\hline
\hline
% LBA fitting: 10-2000
$N_\textrm{total}$ & Total number of samples in histogram & $8.0\times10^{11}$ & $5.4\times10^{11}$ \\
$\sigma$ & Rayleigh mode (assumed to be SEFD/$\sqrt{2\Delta t \Delta \nu}$, & $770$ Jy & $77$ Jy \\
 & where SEFD is the System Equivalent Flux Density) & & \\
\hline
\multicolumn{4}{l}{\textit{Estimators for power-law distribution parameters}} \\
\hline
$\alpha$ & Exponent of power law in RFI distribution      & $-1.62$ & $-1.53$ \\ %-1.616 & -1.523
$SE(\alpha)$ & Standard error of $\alpha$ & $2.8 \times 10^{-3}$ & $6.9 \times 10^{-4}$ \\
$\alpha_H$ & Hill estimator for power-law exponent & $-1.53$ & $-1.53$ \\
$SE(\alpha_H)$ & Standard error of $\alpha_H$ & $8.9 \times 10^{-6}$ & $1.0\times10^{-5}$ \\
$\epsilon_{\alpha}$ & Sampled estimate of standard deviation of $\alpha$ & $6.1 \times 10^{-2}$ & $1.2 \times 10^{-2}$ \\
$\beta$ & Scaling factor of power law with exponent $\alpha$ & $4.0 \times 10^{17}$ & $3.4\times 10^{15}$ \\
$\eta$ & Radiation fall-off speed for $\alpha$ ($\eta=2$ is free space) & 3.23 & 3.77 \\
\hline
\multicolumn{4}{l}{\textit{Limits}} \\
\hline
$S_L$ & Constraint on lower fall-off point of power law & $21$ mJy & $6.2$ mJy \\ % $10^{-6.42}$ & $10^{-4.31}$
$\tilde{S}_L$ & As $S_L$, but assuming $Ig/r^2\sim$ uniform & $47$ mJy & $14$ mJy \\ % $10^{-6.08}$ & $10^{-3.96}$
$S_d$ & Expected lowest apparent level of RFI detected & $26$ Jy & $5.7$ Jy\\ % $2.3 \times 10^{-2}$ & $4.5 \times 10^{-2}$
% $S_U$ & Upper fall-off point & n.a. & $10^{2.62}$ \\
%$\epsilon_R$ & Error in Rayleigh part & TODO & \\
$E(S_R)$ & Apparent RFI flux density & $2,700$ Jy & $140$ Jy \\
$E(S_\textrm{leak})$& Residual apparent RFI flux density after excision & $484$--$496$ mJy & $167$--$171$ mJy \\
 & Same as above, but by assuming 10\% occupancy & $384$ mJy & $120$ mJy \\ % 97--123 mJy
REFD& RFI equivalent flux density & $18.9$--$19.3$ Jy & $6.5$--$6.7$ Jy \\
\hline
\multicolumn{4}{l}{\textit{Average station temperatures}} \\
\hline
$T_\textrm{sys}$ & System temperature (in clean bands)  & 5,000 K & 640 K \\
$T_\textrm{R}$ & Temperature of RFI & 17,000 K & 1,200 K \\
$T_\textrm{leak}$ & Temperature of undetected RFI & 3.2 K & 1.4 K \\
\hline
\hline
\end{tabular}
\end{minipage}
\end{table*}

@@

\citet{lf-interference-temperature-rogers} TODO

>>>>

Typical radio observations record a few hours of data, and the results are integrated. In these cases, excising only the interference that is apparent and thus above the noise often suffices, i.e., the observation can still reach the thermal noise limit of the instrument. A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, weak interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be persistently present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for the 21-cm Epoch of Reionisation (EoR) experiments, because they involve long integration times. Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}. Ground-based Cosmic Microwave Background (CMB) experiments are another class of experiments involving long integration times (e.g., \citealt{ska-cmb-subrahmanyan}). For these experiments, it is important to know the possible effect of low-level interference on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, van Haarlem et al. 2013, \textit{submitted}).
% TODO correct reference if van Haarlem et al is published
\nocite{lofar-overview-2013}

@@

\begin{figure}
\begin{center}\hspace{-5mm}\includegraphics[width=8cm]{img/plot-rayleigh-and-rfi-trimmed}
\caption{The Rayleigh and power-law distributions in a log-log plot. The power-law distribution (Eq.~\eqref{eq:two-dimensional-distribution}) has a constant slope of -2 over the range it is defined. The slope of the Rayleigh distribution in the limit of the origin is 1. Its maximum occurs where the amplitude value equals its mode $\sigma$, which is 1 in this example. For higher amplitudes, its slope decreases exponentially.}
\label{fig:rayleigh-and-rfi-distributions}
\end{center}
\end{figure}

@@

The stochastic nature of $I$, that is caused by the spread of transmitters with different intrinsic strengths, might also have an effect on the $\log N$ -- $\log S$ histograms. It is unlikely that $I$ follows a uniform or Gaussian distribution, because the distribution will contain few strong transmitters (such as radio stations) and many weak transmitters (such as remote controls). Therefore, variable $I$ might follow a power-law distribution by itself. It is likely that strong transmitters transmit more on average, and therefore contaminate more samples as well. High-power transmitters, such as radio stations, have a typical transmitting power in the order of 100~kW. Low-power transmitters, such as remote controls, transmit with an order of 100~mW or even less. Therefore, these devices have a spread of around 6 orders of magnitudes in power. As long as the exponent of the power law of $I$ is less steep (i.e., less negative) compared to the power law caused by the spatial distribution, the spatial distribution will dominate the histogram at high amplitudes. With a spatial $-1.5$ power law and the given transmitting powers, the low-power transmitters should contaminate a factor of $10^9$ more samples compared to the high-power transmitters to dominate the high-amplitude distribution, which is unlikely. Therefore, it is likely that the spatial distribution will dominate the power law in the histogram. Otherwise, a turn-over point should be visible in the histogram.

@@

\begin{table*}
\centering
\begin{minipage}{12cm}
\caption{Estimated distribution quantities per data set. }\label{table:dist-data-quantities}
\begin{tabular}{@{}llrr@{}}
\textbf{Symbol} & \textbf{Description} & \textbf{LBA set}& \textbf{HBA set} \\
\hline
\hline
% LBA fitting: 10-2000
$N_\textrm{total}$ & Total number of samples in histogram & $8.0\times10^{11}$ & $5.4\times10^{11}$ \\
$\sigma$ & Rayleigh mode (assumed to be SEFD/$\sqrt{2\Delta t \Delta \nu}$, & $770$ Jy & $77$ Jy \\
 & where SEFD is the System Equivalent Flux Density) & & \\
\hline
\multicolumn{4}{l}{\textit{Estimators for power-law distribution parameters}} \\
\hline
$\alpha$ & Exponent of power law in RFI distribution      & $-1.62$ & $-1.53$ \\ %-1.616 & -1.523
$SE(\alpha)$ & Standard error of $\alpha$ & $2.8 \times 10^{-3}$ & $6.9 \times 10^{-4}$ \\
$\alpha_H$ & Hill estimator for power-law exponent & $-1.53$ & $-1.53$ \\
$SE(\alpha_H)$ & Standard error of $\alpha_H$ & $8.9 \times 10^{-6}$ & $1.0\times10^{-5}$ \\
$\epsilon_{\alpha}$ & Sampled estimate of standard deviation of $\alpha$ & $6.1 \times 10^{-2}$ & $1.2 \times 10^{-2}$ \\
$\beta$ & Scaling factor of power law with exponent $\alpha$ & $4.0 \times 10^{17}$ & $3.4\times 10^{15}$ \\
$\eta$ & Radiation fall-off speed for $\alpha$ ($\eta=2$ is free space) & 3.23 & 3.77 \\
\hline
\multicolumn{4}{l}{\textit{Limits}} \\
\hline
$S_L$ & Constraint on lower fall-off point of power law & $21$ mJy & $6.2$ mJy \\ % $10^{-6.42}$ & $10^{-4.31}$
$\tilde{S}_L$ & As $S_L$, but assuming $Ig/r^2\sim$ uniform & $47$ mJy & $14$ mJy \\ % $10^{-6.08}$ & $10^{-3.96}$
$S_d$ & Expected lowest apparent level of RFI detected & $26$ Jy & $5.7$ Jy\\ % $2.3 \times 10^{-2}$ & $4.5 \times 10^{-2}$
% $S_U$ & Upper fall-off point & n.a. & $10^{2.62}$ \\
%$\epsilon_R$ & Error in Rayleigh part & TODO & \\
$E(S_R)$ & Apparent RFI flux density & $2,700$ Jy & $140$ Jy \\
$E(S_\textrm{leak})$& Residual apparent RFI flux density after excision & $484$--$496$ mJy & $167$--$171$ mJy \\
 & Same as above, but by assuming 10\% occupancy & $384$ mJy & $120$ mJy \\ % 97--123 mJy
REFD& RFI equivalent flux density & $18.9$--$19.3$ Jy & $6.5$--$6.7$ Jy \\
\hline
\multicolumn{4}{l}{\textit{Average station temperatures}} \\
\hline
$T_\textrm{sys}$ & System temperature (in clean bands)  & 5,000 K & 640 K \\
$T_\textrm{R}$ & RFI Temperature & 17,000 K & 1,200 K \\
$T_\textrm{leak}$ & Temperature of undetected RFI & 3.2 K & 1.4 K \\
\hline
\hline
\end{tabular}
\end{minipage}
\end{table*}

@@

The apparent RFI flux densities can be converted to a RFI station temperature that excludes the system noise and sky noise components. If we use a station efficiency factor $\eta_\textrm{st}=1$ and effective areas LBA $A_\textrm{eff}=398$ and HBA $A_\textrm{eff}=512$ with again 100\% RFI occupancy, our models lead to RFI temperatures of 17,000~K and 1,200~K for respectively the LBA and the HBA. These are relatively high compared to for example the survey by \citet{lf-interference-temperature-rogers}, who report that on two different sites, 20\% and 27\% of the spectrum has a temperature above 450~K in the range of 50--1500~MHz. However, our post-detection RFI station temperatures, which arise from the residual apparent RFI flux density estimates, are 3.2~K and 1.4~K for the LBA and HBA respectively. Due to LOFAR's high resolution and accurate flagging strategy, this is achieved by flagging a relatively small data percentage of 1.8 (LBA) and 3.2\% (HBA).
